---
title: "Prediction Assignment Writeup"
author: "Rodrigo Morais"
date: "11/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(rpart)
library(rpart.plot)
library(rattle)
library(caret)
library(randomForest)
set.seed(1104)
```

## Introduction

This is a requirement in the Practical Machine Learning course by Johns Hopkins University. The goal of this assignment is to predict the manner of the subject (6 participants) performed exercises. The data was obtained from devices such as Jawbone Up, Nike FuelBand, and Fitbit. In order to predict the manner of the subject did the exercises we will be using the decision tree and random forest methods to determine the best prediction with the highest accuracy possible.

## Data

The training data for this project is available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data is available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

## Analysis 

### Data acquisition and transformation
```{r cache=TRUE}
# downloading and loading data
test_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"

testing <- read.csv(url(test_url), na.strings = c("", "NA"))
training <- read.csv(url(train_url), na.strings = c("", "NA"))

```

```{r}
dim(testing)
dim(training)

```

Cross-validation will be performed by spliting the training dataset to a new one with 70% of the observations. The models for prediction will be built using this dataset. The testing dataset will have 30% of the observations, and will be used to evalueate the accuracy of the prediction models.

```{r}
partdata <- createDataPartition(y = training$classe, p = 0.7, list = FALSE)
traindata <- training[partdata,]
testdata <- training[-partdata,]
testdata$classe <- as.factor(testdata$classe)

dim(traindata)
dim(testdata)
```
```{r}
nearzero <- nearZeroVar(traindata)
traindata <- traindata[, -nearzero]
testdata  <- testdata[, -nearzero]

# removing nomenclature columns and variable with too many NA values
traindata <- traindata[, -(1:5)]
testdata <- testdata[, -(1:5)]
nona <- sapply(traindata, function(x) mean(is.na(x))) > 0.90
traindata <- traindata[, nona == FALSE]
testdata  <- testdata[, nona == FALSE]
testdata$classe <- as.factor(testdata$classe)
```

## Prediction Models

### Decision Tree
```{r cache=TRUE}
# traindata dt
dtmod <- rpart(classe ~ ., data = traindata, method = "class")
fancyRpartPlot(dtmod)

# testdata prediction
pred_test <- predict(dtmod, newdata = testdata, type = "class")
conf_test <- confusionMatrix(pred_test, testdata$classe)
conf_test

# plot
plot(conf_test$table, col = conf_test$byClass, main = paste("Decision Tree Accuracy =", round(conf_test$overall['Accuracy'], 3)))

```

### Random Forest
```{r cache=TRUE}
# traindata rf
randof <- trainControl(method = "cv", number = 3, verboseIter = FALSE)
randomodfit <- train(classe ~ ., data = traindata, method = "rf", trControl = randof)
randomodfit$finalModel

# testdata prediction
pred_train <- predict(randomodfit, newdata = testdata)
conf_train <- confusionMatrix(pred_train, testdata$classe)
conf_train

# plot
plot(conf_train$table, col = conf_train$byClass, main = paste("Random Forest Accuracy = ", round(conf_train$overall['Accuracy'], 3)))
```

## Using Randon Forest to find prediction cases
```{r}
pred_rf <- predict(randomodfit, newdata = testing)
pred_rf

```
